The Nuclear Forensics Problem and Statistical Methods: \\ Evaluating Machine Learning for Spent Fuel Prediction

### 2017-05-18 paper.tex
updates

Workflow for determining SNF quantities of interest: measure material, use
isotope content and/or isotope ratios to determine things like reactor type,
fuel type and enrichment at beginning of irradiation, cooling time, burnup.

After the material characteristics are measured, they are matched in a
forensics database that includes some or all this information for
pre-existing/pre-measured SNF. These databases are kept by individual
countries, and a given database will have widely varying uncertainty depending
where the material was measured as well as missing data in some fields.
Therefore, matching can be difficult.

(Classification, Characterization, Interpretation (Analysis), Reconstruction
(Attribution) - from the New Nuclear Forensics book) (Char methods to get
isotopic ratios or use S/ML, Interp examples)

To accomplish this, (we want to see if) it is possible to limit material
measurements to rapid-result nondestructive analysis, such as gamma ray
spectroscopy. Gamma spec gives less information at a higher uncertainty than
the near perfect results of some destructive mass spec techniques, like TIMS.
Additionally, within gamma spec techniques (field vs lab), uncertainties can
also vary.

(Something on goal to get as much info as possible within 24 h of interception
or device detonation).

@@

Given imperfect data with varying amounts of uncertainty as well as the
required comparison to imperfect databases, many have begun considering
artificial intelligence approaches to nuclear forensics problems, such as
implementing searching algorithms for database comparison and machine learning
for determining spent fuel characteristics (cite all). A variety of statistical
and machine learning tools have been used to both classify spent fuel (reactor
type, fuel type) and predict values such as burnup, initial enrichment, or
cooling time (regression).

Add some real ML background here

@@

This work begins by replicating the training and test sets used in ref (cite
Dayman). The study does evaluate the imapact of randomly introduced error of
varying amounts on the ability of the algorithms to correctly predict the
burnup and reactor type. It also varies the amount of nuclides given as
training to the algorithms (a full list, a short list determined by PCA, and
fission products only). However, machine learning algorithms are heavily
dependent on the inputs and parameters given to them, such as training set
sizes, learning rates, etc. Additionally, it does not consider more advanced
aglorithms. This paper adds in the comparison of the simpler algorithms to
neural nets. It also presents learning curves for each algorithm to investigate
the effect of the training set size on prediction error. For the classification
of reactor type, ROC curves are also presented to evaluate prediction and
generalization strength. For the regression of burnup, ??? pairwise t test?.
Thus, I will first investigate these algorithms with no error introduced, but
will extend the validation from a predetermined test set to be randomly
partitioned cross validation.  And the next study will introduce error by
limiting the nuclides to only those that can be measured with a gamma
spectrometer (future work).

>>>>

This study focuses on non-detonated materials, specifically, spent nuclear
fuel. It is important to determine if some intercepted material is from a
commercial fuel cycle or if it is meant for weapons production (and where the
material was obtained from).  Workflow for determining SNF quantities of
interest: measure material, use isotope content and/or isotope ratios to
determine things like reactor type, fuel type and enrichment at beginning of
irradiation, cooling time, burnup.  (Classification, Characterization,
Interpretation (Analysis), Reconstruction (Attribution) - from the New Nuclear
Forensics book) (Char methods to get isotopic ratios or use S/ML, Interp
examples) After the material characteristics are measured, they are matched in
a forensics database that includes some or all this information for
pre-existing/pre-measured SNF. These databases are kept by individual
countries, and a given database will have widely varying uncertainty depending
where the material was measured as well as missing data in some fields.
Therefore, matching can be difficult.

So a lofty goal for the forensics community would be to develop methods that
provide instantaneous information that is reliable enough to guide an
investigation (e.g., within 24 hours). Fast, accurate measurements to provide
isotopic ratios to calculate the above-mentioned fuel parameters of interest
would provide this. To accomplish/mimic this, we wil limit material
measurements to rapid-result nondestructive analysis, such as gamma ray
spectroscopy. Gamma spec gives less information at a higher uncertainty than
the near perfect results of some destructive mass spec techniques, like TIMS.
Additionally, within gamma spec techniques (field vs lab), uncertainties can
also vary. However, using a well-trained machine-learned model may be able to
overcome these inherent issues with gamma spec. The current and future work of
this study is designed with this in mind.

@@

Given imperfect data with varying amounts of uncertainty as well as the
required comparison to imperfect databases, many have begun considering
artificial intelligence approaches to nuclear forensics problems, such as
implementing searching algorithms for database comparison and machine learning
for determining spent fuel characteristics (cite all). A variety of statistical
and machine learning tools have been used to both classify spent fuel (reactor
type, fuel type) and provide regression values such as burnup, initial
enrichment, or cooling time.

In order to validate my own work, I am choosing some machine learning
approaches based on previous work (cite) while also extending to more complex
algorithms. Thus, this paper investigates four algorithms: nearest neighbors
using two different distance metrics, ridge regression, and a neural network.
These algorithms can all be modified to do either classification or regression.

In this paper, run through em all quickly, but in prelim make each into it's
own section.

@@

This work begins by replicating the training and test sets used in ref (cite
Dayman). Replicate table of training instances perhaps. Discuss how this covers
both short and long fuel cycles and their relation to weapons versus commercial
use.

The study does evaluate the imapact of randomly introduced error of
varying amounts on the ability of the algorithms to correctly predict the
burnup and reactor type. It also varies the amount of nuclides given as
training to the algorithms (a full list, a short list determined by PCA, and
fission products only). However, machine learning algorithms are heavily
dependent on the inputs and parameters given to them, such as training set
sizes, learning rates, etc. Additionally, it does not consider more advanced
aglorithms. This paper adds in the comparison of the simpler algorithms to
neural nets. It also presents learning curves for each algorithm to investigate
the effect of the training set size on prediction error. For the classification
of reactor type, ROC curves are also presented to evaluate prediction and
generalization strength. For the regression of burnup, ??? pairwise t test?.
Thus, I will first investigate these algorithms with no error introduced, but
will extend the validation from a predetermined test set to be randomly
partitioned cross validation.  And the next study will introduce error by
limiting the nuclides to only those that can be measured with a gamma
spectrometer (future work).
